Advent Of Code 2020

### 2020-12-06 AdvendOfCode2020/draft_post.Rmd
day 3 part 1

```{r}
# separete and str_count come in handy for this
library(tidyverse)

dplyr:

# read the data as data.
password.check <- read.csv("./day02_input.txt", header = F) %>% 
  set_names(c("input")) %>% 
  separate(input, c("policy","password"), sep=": ", remove=F) %>% 
  separate(policy, c("policy.range","policy.letter"), sep=" ") %>%
  separate(policy.range, c("min","max"), sep = "-", convert = T) %>% 
  mutate( count.letter = str_count(password, policy.letter)) %>% 
  mutate( valid.password = count.letter >= min & count.letter <= max )

# how many password are valid?
sum(password.check$valid.password)

@@

>>>>

```{r}
# tidyr::separate() and stringr::str_count() come in handy for this
library(tidyverse)

# read the input data as data frame with a column named 'input'
password.check <- read.csv("./day02_input.txt", header = F) %>% 
  set_names(c("input")) %>% 
  # split the line into the password and the policy columns
  separate(input, c("policy","password"), sep=": ", remove=F) %>% 
  # split the policy into range values and the letter to check
  separate(policy, c("pol.range.min", "pol.range.max","pol.letter"),
           sep="-| ", convert=T) %>% 
  # count the letter in the password and check the policy range
  mutate( count.letter = str_count(password, pol.letter),  
          is.valid = count.letter >= pol.range.min &
                     count.letter <= pol.range.max )

# let's see what we got
head(password.check)

# how many password are valid?
sum(password.check$is.valid)

```
### Part Two

While it appears you validated the passwords correctly, they don't seem to be what the Official Toboggan Corporate Authentication System is expecting.

The shopkeeper suddenly realizes that he just accidentally explained the password policy rules from his old job at the sled rental place down the street! The Official Toboggan Corporate Policy actually works a little differently.

Each policy actually describes two positions in the password, where 1 means the first character, 2 means the second character, and so on. (Be careful; Toboggan Corporate Policies have no concept of "index zero"!) Exactly one of these positions must contain the given letter. Other occurrences of the letter are irrelevant for the purposes of policy enforcement.

Given the same example list from above:

1-3 a: abcde is valid: position 1 contains a and position 3 does not.
1-3 b: cdefg is invalid: neither position 1 nor position 3 contains b.
2-9 c: ccccccccc is invalid: both position 2 and position 9 contain c.

How many passwords are valid according to the new interpretation of the policies?

```{r}
# tidyr::separate() and stringr::str_count() come in handy for this
library(tidyverse)

# read the input data as data frame with a column named 'input'
password.check <- read.csv("./day02_input.txt", header = F) %>% 
  set_names(c("input")) %>% 
  # split the line into the password and the policy columns
  separate(input, c("policy","password"), sep=": ", remove=F) %>% 
  # split the policy into range values and the letter to check
  separate(policy, c("pol.range.min", "pol.range.max","pol.letter"),
           sep="-| ", convert=T) %>% 
  # part 1 
  # count the letter in the password and check the policy range
  mutate( count.letter = str_count(password, pol.letter),  
          is.valid.part1 = count.letter >= pol.range.min &
                           count.letter <= pol.range.max ) %>% 
  # part 2
  # get the letters in positions 'min' and 'max"' and "XOR" check
  mutate(
    letter.at.pos1 = str_sub(password, pol.range.min, pol.range.min),
    letter.at.pos2 = str_sub(password, pol.range.max, pol.range.max),
    is.valid.part2 = xor(letter.at.pos1==pol.letter, letter.at.pos2==pol.letter)
  )

# part one: how many password are valid (repeat letter policy)?
sum(password.check$is.valid.part1)

# part tow: how many password are valid (letter at position xor policy?
sum(password.check$is.valid.part2)

```
## Day 3: Toboggan Trajectory

### Part 1

With the toboggan login problems resolved, you set off toward the airport. While travel by toboggan might be easy, it's certainly not safe: there's very minimal steering and the area is covered in trees. You'll need to see which angles will take you near the fewest trees.

Due to the local geology, trees in this area only grow on exact integer coordinates in a grid. You make a map (your puzzle input) of the open squares (.) and trees (#) you can see. For example:

```
..##.......
#...#...#..
.#....#..#.
..#.#...#.#
.#...##..#.
..#.##.....
.#.#.#....#
.#........#
#.##...#...
#...##....#
.#..#...#.#
```

These aren't the only trees, though; due to something you read about once involving arboreal genetics and biome stability, the same pattern repeats to the right many times:

```
..##.........##.........##.........##.........##.........##.......  --->
#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..
.#....#..#..#....#..#..#....#..#..#....#..#..#....#..#..#....#..#.
..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#
.#...##..#..#...##..#..#...##..#..#...##..#..#...##..#..#...##..#.
..#.##.......#.##.......#.##.......#.##.......#.##.......#.##.....  --->
.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#
.#........#.#........#.#........#.#........#.#........#.#........#
#.##...#...#.##...#...#.##...#...#.##...#...#.##...#...#.##...#...
#...##....##...##....##...##....##...##....##...##....##...##....#
.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#.#..#...#.#  --->
```

You start on the open square (.) in the top-left corner and need to reach the bottom (below the bottom-most row on your map).

The toboggan can only follow a few specific slopes (you opted for a cheaper model that prefers rational numbers); start by counting all the trees you would encounter for the slope right 3, down 1:

From your starting position at the top-left, check the position that is right 3 and down 1. Then, check the position that is right 3 and down 1 from there, and so on until you go past the bottom of the map.

The locations you'd check in the above example are marked here with O where there was an open square and X where there was a tree:

```
..##.........##.........##.........##.........##.........##.......  --->
#..O#...#..#...#...#..#...#...#..#...#...#..#...#...#..#...#...#..
.#....X..#..#....#..#..#....#..#..#....#..#..#....#..#..#....#..#.
..#.#...#O#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#..#.#...#.#
.#...##..#..X...##..#..#...##..#..#...##..#..#...##..#..#...##..#.
..#.##.......#.X#.......#.##.......#.##.......#.##.......#.##.....  --->
.#.#.#....#.#.#.#.O..#.#.#.#....#.#.#.#....#.#.#.#....#.#.#.#....#
.#........#.#........X.#........#.#........#.#........#.#........#
#.##...#...#.##...#...#.X#...#...#.##...#...#.##...#...#.##...#...
#...##....##...##....##...#X....##...##....##...##....##...##....#
.#..#...#.#.#..#...#.#.#..#...X.#.#..#...#.#.#..#...#.#.#..#...#.#  --->
```

In this example, traversing the map using this slope would cause you to encounter 7 trees.

Starting at the top-left corner of your map and following a slope of right 3 and down 1, how many trees would you encounter?

```{r}

# read the files as vector of string
input <- read.csv("./day03_input.txt", header = F)[,1]

# create a "char" matrix of "." and "#"
char_map <- strsplit(input, "") %>% 
  unlist() %>% 
  matrix(nrow = length(input), byrow = T)

# convert in to 0 and 1 (1 for the trees) 
int_map <- 1*(char_map == "#")

# lets subset the matrix with the trajectory

# the hill trajectory: 3 to the left and 1 to the bottom
shifts <- c(
  0,1,
  0,2,
  0,3,
  1,3
) %>% matrix(nrow=4, byrow=T)

# starting from top left
base_pos <- c(1,1)
trajectory <- base_pos

# until the last row
for(i in 1:(nrow(int_map)-1)){
  steps <- rep(base_pos,4) %>% 
    matrix(nrow = 4, byrow = T) +
    shifts

  base_pos <- steps[4,]

  trajectory <- rbind(trajectory, base_pos)
}

# the columns repeat adjusting
y_coords <- trajectory[,2] %% ncol(int_map)
y_coords[y_coords==0] <- ncol(int_map)

trajectory[,2] <- y_coords

sum(int_map[trajectory])

@@

### Part 2

Time to check the rest of the slopes - you need to minimize the probability of a sudden arboreal stop, after all.

Determine the number of trees you would encounter if, for each of the following slopes, you start at the top-left corner and traverse the map all the way to the bottom:

Right 1, down 1.
Right 3, down 1. (This is the slope you already checked.)
Right 5, down 1.
Right 7, down 1.
Right 1, down 2.
In the above example, these slopes would find 2, 7, 3, 4, and 2 tree(s) respectively; multiplied together, these produce the answer 336.

What do you get if you multiply together the number of trees encountered on each of the listed slopes?

```{r}
# read the files as vector of string
input <- read.csv("./day03_input.txt", header = F)[,1]

# create a "char" matrix of "." and "#"
char_map <- strsplit(input, "") %>% 
  unlist() %>% 
  matrix(nrow = length(input), byrow = T)

# convert in to 0 and 1 (1 for the trees) 
int_map <- 1*(char_map == "#")

# lets subset the matrix with the trajectory

# the hill trajectory: 3 to the left and 1 to the bottom
shifts <- c(
  0,1,
  0,2,
  0,3,
  1,3
) %>% matrix(nrow=4, byrow=T)

# starting from top left
base_pos <- c(1,1)
trajectory <- base_pos

# until the last row
for(i in 1:(nrow(int_map)-1)){
  steps <- rep(base_pos,4) %>% 
    matrix(nrow = 4, byrow = T) +
    shifts

  base_pos <- steps[4,]

  trajectory <- rbind(trajectory, base_pos)
}

# the columns repeat adjusting
y_coords <- trajectory[,2] %% ncol(int_map)
y_coords[y_coords==0] <- ncol(int_map)

trajectory[,2] <- y_coords

sum(int_map[trajectory])
